This script reads in a data file containing driving distances and fares between every pair of stations served by Bay Area Rapid Transit, as of January 1, 2013. Airport = 1 if and only if San Francisco International Airport is the To or From Station. I assume Fare is symmetric, so that traveling from Station A to Station B costs the same as B to A and has the same distance. SanMateo == 1 iff the trip is To or From Daly City, Colma, South San Francisco, San Bruno or Millbrae. SFO airport might be in San Mateo county, but I already have a surcharge built in for SFO trips.  
Written by Timothy Sweetser, last update April 2014.
https://sites.google.com/site/tsweetser

Read in raw data:
```{r}
setwd('//volumes/no name/BART project')
fare_attributes <- read.csv("google_transit_2/fare_attributes.txt", header=T)
fare_rules <- read.csv("google_transit_2/fare_rules.txt", header=T)
fares <- merge(fare_attributes, fare_rules, by = "fare_id")
```
Since there are 44 stations, Fares has $1936 = 44^2$ rows.  
Read in miles between pairs of stations:
```{r}
distances <- read.csv("data/BART/Dijkstra miles.csv", header=T)
distances$key <- paste(distances$From, distances$To, sep="")
fares$key <- paste(fares$origin_id, fares$destination_id, sep="")
dataset <- merge(fares, distances, by="key")
```
Remove trips starting and beginning at the same station.
```{r}
BART.raw <- BART.raw[BART.raw$From != BART.raw$To, ]
```
Airport and San Mateo dummy variables
```{r}
BART.raw$Airport <- as.numeric(BART.raw$To == 'SFIA' | 
                                 BART.raw$From == 'SFIA')
stations.SM <- c('Daly City', 'Colma', 'South SF', 
                       'San Bruno', 'Millbrae')
BART.raw$SanMateo <- as.numeric(BART.raw$To %in% stations.SM |
                      BART.raw$From %in% stations.SM)
```
Include each trip (one pair of stations) only once
```{r}
keepers <- NULL
for (i in 1: nrow(BART.raw)) {
  sister.row <- intersect(which(BART.raw$To == BART.raw$From[i]),
                      which(BART.raw$From == BART.raw$To[i]))
  keepers <- c(keepers,min(i,sister.row,na.rm=T))
}
BART.fares <- BART.raw[unique(keepers), ]
rm(list=c('i', 'keepers', 'sister.row')) # cleanup
```
TransBay dummy variable 
```{r}
bayside <- read.csv("data/BART/stations side of bay.csv")
BART.fares <- merge(x=BART.fares, y=bayside, by.x='From', 
                    by.y='stations')
names(BART.fares)[ncol(BART.fares)] <- "From_Side"
BART.fares <- merge(x=BART.fares, y=bayside, by.x='To', 
                    by.y='stations')
names(BART.fares)[ncol(BART.fares)] <- "To_Side"
BART.fares$TransBay <- as.numeric(BART.fares$To_Side != 
                                    BART.fares$From_Side)
```
Now, fit a linear regression model
```{r}
mat <- lm(Fare ~ 0 + Miles + Airport*TransBay + SanMateo*TransBay +
            Miles:TransBay + Miles:Airport + Miles:SanMateo, 
          data=BART.fares, offset=rep(1.75,nrow(BART.fares)))
summary(mat)
```
None of these interaction terms are significant at 5%: Miles:TransBay, Miles2:Airport, Miles2:SanMateo. This means the per-mile rate is the same throughout the system. The Millbrae-SFO trip has an extremely large, negative residual. How much do the coefficients change by excluding it?
```{r}
summary(lm(Fare ~ 0 + Miles + Airport*TransBay + SanMateo*TransBay, 
           data=BART.fares, offset=rep(1.75,nrow(BART.fares)),
           subset=c(rep(TRUE, 831), FALSE, rep(TRUE, 946-832))))
```
Aiport increases from 4.43 to 4.70.  
Airport:TransBay decreases from -0.73 to -1.00.  
SanMateo increased from 0.80 to 0.84.  

Compare SFO - Millbrae (row 832) to SFO - San Bruno (811)
similar miles, but $3 fare difference!

Use this mask to color-code trips
```{r}
transbay.mask <- BART.fares$TransBay == 1 
airport.mask <- BART.fares$Airport == 1

plot(x=BART.fares$Miles2, y=mat$residuals)
```

Outliers: 911 San Bruno - SFO, 921 SFO - South SF
Also SFO - Millbrae: few miles, high fare  

Can the model figure out the intercept on its own?
```{r}
m.int <- lm(Fare ~ Miles + Airport + TransBay, data=BART.fares) 
``` 
Intercept of 1.64. pretty close

First Graph: just the data
```{r}
# non airport, non transbay
plot(BART.fares[!transbay.mask & !airport.mask, 'Miles'],
     BART.fares[!transbay.mask & !airport.mask, 'Fare'],
     col="blue", xlab = "Driving Distance Between Stations (Miles)", 
     ylab= "Fare (Dollars)", xlim=c(0, 52), ylim=c(0, 12),
     main="BART Fare vs. Distance between Stations", pch=0)
# non-airport, transbay
points(BART.fares[transbay.mask & !airport.mask, 'Miles'],
       BART.fares[transbay.mask & !airport.mask, 'Fare'],
       col="orange", pch=0)
# airport, non transbay
points(BART.fares[!transbay.mask & airport.mask, 'Miles'],
       BART.fares[!transbay.mask & airport.mask, 'Fare'],
       col="blue", pch=17)
# airport, transbay
points(BART.fares[transbay.mask & airport.mask, 'Miles'],
       BART.fares[transbay.mask & airport.mask, 'Fare'],
       col="orange", pch=17)
 # blue = non-transbay, orange=transbay
# pch0=square = non-airport. pch17=triangle=airport
legend(x="bottomright", legend=c('No Tube No SFO', 
                                 'Cross Bay noSFO',
                                 'SFO-Peninsula',
                                 'SFO-East Bay'), 
       col=rep(c("blue", "orange"), 2), pch=c(0, 0, 17, 17))
```
Now add regression lines.
```{r}
plot(BART.fares[!transbay.mask & !airport.mask, 'Miles'],
     BART.fares[!transbay.mask & !airport.mask, 'Fare'],
     col="blue", xlab = "Driving Distance Between Stations (Miles)", 
     ylab= "Fare (Dollars)", xlim=c(0, 52), ylim=c(0, 12),
     main="BART Fare vs. Distance between Stations", pch=0)
# non-airport, transbay
points(BART.fares[transbay.mask & !airport.mask, 'Miles'],
       BART.fares[transbay.mask & !airport.mask, 'Fare'],
       col="orange", pch=0)
# airport, non transbay
points(BART.fares[!transbay.mask & airport.mask, 'Miles'],
       BART.fares[!transbay.mask & airport.mask, 'Fare'],
       col="blue", pch=17)
# airport, transbay
points(BART.fares[transbay.mask & airport.mask, 'Miles'],
       BART.fares[transbay.mask & airport.mask, 'Fare'],
       col="orange", pch=17)
# blue = non-transbay, orange=transbay
# pch0=square = non-airport. pch17=triangle=airport
legend(x="bottomright", legend=c('No Tube No SFO', 
                                 'Cross Bay noSFO',
                                 'SFO-Peninsula',
                                 'SFO-East Bay'), 
       col=rep(c("blue", "orange"), 2), pch=c(0, 0, 17, 17))

abline(a=1.75, b=coef(mat)[1], lwd=2, col="blue") # no tube, no airport
abline(a=coef(mat)[3] + 1.75, b=coef(mat)[1], lwd=2, 
       col="orange") #tube, no airport
abline(a=coef(mat)[2] + 1.75, b=coef(mat)[1], lwd=2,
       col="blue") #no tube, airport
abline(a=sum(coef(mat)[2:4]) + 1.75, b=coef(mat)[1], lwd=2,
       col="orange") #tube, airport
```
SFO outlier is Millbrae-SFO trip. Fare = 4.05

Residuals
Airport-only model
```{r}
par(mfrow=c(2, 2))
plot(mat)
par(mfrow=c(1, 1))


plot(y=mat$residuals, x=BART.fares$Miles,
      xlab = "Driving Distance Between Stations (Miles)",
     ylab="Residual", main="Residuals of Model")
```


Look at mean residual by station.
```{r}
boxplot(sapply(split(c(mat$residual, mat$residual), 
             c(BART.fares$To, BART.fares$From)), mean))
lapply(split(c(mat$residual, mat$residual), 
             c(BART.fares$To, BART.fares$From)), length)
```
Millbrae has an average residual of +52 cents!