---
title: "BART Fare Formula"
author: "Timothy Sweetser"
date: "January 7, 2015"
output: html_document
---

This script reads in a data file containing driving distances and fares between every pair of stations served by Bay Area Rapid Transit, as of January 1, 2013. 
* Self and Airport are indicator variables  
* Self = 1 if and only if From Station = To Station  
* Airport = 1 if and only if San Francisco International Airport is the To or From Station  
* I assume Fare is symmetric, so that traveling from Station A to Station B costs the same as B to A  
* SanMateo == 1 iff the trip is To or From Daly City, Colma, South San Francisco, San Bruno or Millbrae  
* SFO airport might be in San Mateo county, but I already have a surcharge built in for SFO trips  
* http://sites.google.com/site/tsweetser  

```{r}
setwd('/Users/timothysweetser/Box Sync/BART Project')
BART.raw <- read.csv("data/bart/BART all possible trips.csv")
source("distance correct.R")
BART.raw <- dist_correct(BART.raw)

# include each trip (one pair of stations) only once
keepers <- NULL
for (i in 1: nrow(BART.raw)) {
  sister.row <- intersect(which(BART.raw$To == BART.raw$From[i]),
                      which(BART.raw$From == BART.raw$To[i]))
  keepers <- c(keepers,min(i,sister.row,na.rm=T))
}
rm(i)
BART.fares <- BART.raw[unique(keepers), ]

# fit a linear regression model
mat <- lm(Fare ~ 0 + Miles2 + Airport*TransBay + SanMateo*TransBay, 
          # Airport:SanMateo is all 1's
          data=BART.fares,
          offset=rep(1.75,nrow(BART.fares)))
```

None of these interaction terms are significant at $\alpha=0.05$ significance: Miles2:TransBay, Miles2:Airport, Miles2:SanMateo. This means the per-mile rate is the same throughout the system.  
The Millbrae-SFO trip has an extremely large, negative residual. How much do the coefficients change by excluding it?
```{r}
summary(lm(Fare ~ 0 + Miles2 + Airport*TransBay + SanMateo*TransBay, 
           data=BART.fares, offset=rep(1.75,nrow(BART.fares)),
           subset=BART.fares$Station.Pair.ID != 1110))
```
* Aiport increases from 4.84 to 5.09  
* Airport:TransBay decreases from -0.84 to -1.09  
* Use this mask to color-code trips
```{r}
transbay.mask <- BART.fares$TransBay == 1 
airport.mask <- BART.fares$Airport == 1

plot(x=BART.fares$Miles2, y=mat$residuals)
identify(x=BART.fares$Miles2, y=mat$residuals)
```
Outliers include 911 (San Bruno - SFO), and 921 (SFO - South SF), and SFO - Millbrae.  
Can the model figure out the intercept on its own?
```{r}
m.int <- lm(Fare ~ Miles + Airport + TransBay, data=BART.fares) 
```
The fitted intercept of $1.64 is pretty close.  

First Graph: just the data
```{r}
# non airport, non transbay
plot(BART.fares[!transbay.mask & !airport.mask, 'Miles'],
     BART.fares[!transbay.mask & !airport.mask, 'Fare'],
     col="blue", xlab = "Driving Distance Between Stations (Miles)", 
     ylab= "Fare (Dollars)", xlim=c(0, 52), ylim=c(0, 12),
     main="BART Fare vs. Distance between Stations", pch=0)
# non-airport, transbay
points(BART.fares[transbay.mask & !airport.mask, 'Miles'],
       BART.fares[transbay.mask & !airport.mask, 'Fare'],
       col="orange", pch=0)
# airport, non transbay
points(BART.fares[!transbay.mask & airport.mask, 'Miles'],
       BART.fares[!transbay.mask & airport.mask, 'Fare'],
       col="blue", pch=17)
# airport, transbay
points(BART.fares[transbay.mask & airport.mask, 'Miles'],
       BART.fares[transbay.mask & airport.mask, 'Fare'],
       col="orange", pch=17)
 # blue = non-transbay, orange=transbay
# pch0=square = non-airport. pch17=triangle=airport
legend(x="bottomright", legend=c('No Tube No SFO', 
                                 'Cross Bay noSFO',
                                 'SFO-Peninsula',
                                 'SFO-East Bay'), 
       col=rep(c("blue", "orange"), 2), pch=c(0, 0, 17, 17))
```  
Now add regression lines.
```{r}
plot(BART.fares[!transbay.mask & !airport.mask, 'Miles'],
     BART.fares[!transbay.mask & !airport.mask, 'Fare'],
     col="blue", xlab = "Driving Distance Between Stations (Miles)", 
     ylab= "Fare (Dollars)", xlim=c(0, 52), ylim=c(0, 12),
     main="BART Fare vs. Distance between Stations", pch=0)
# non-airport, transbay
points(BART.fares[transbay.mask & !airport.mask, 'Miles'],
       BART.fares[transbay.mask & !airport.mask, 'Fare'],
       col="orange", pch=0)
# airport, non transbay
points(BART.fares[!transbay.mask & airport.mask, 'Miles'],
       BART.fares[!transbay.mask & airport.mask, 'Fare'],
       col="blue", pch=17)
# airport, transbay
points(BART.fares[transbay.mask & airport.mask, 'Miles'],
       BART.fares[transbay.mask & airport.mask, 'Fare'],
       col="orange", pch=17)
# blue = non-transbay, orange=transbay
# pch0=square = non-airport. pch17=triangle=airport
legend(x="bottomright", legend=c('No Tube No SFO', 
                                 'Cross Bay noSFO',
                                 'SFO-Peninsula',
                                 'SFO-East Bay'), 
       col=rep(c("blue", "orange"), 2), pch=c(0, 0, 17, 17))

abline(a=1.75, b=coef(mat)[1], lwd=2, col="blue") # no tube, no airport
abline(a=coef(mat)[3] + 1.75, b=coef(mat)[1], lwd=2, 
       col="orange") #tube, no airport
abline(a=coef(mat)[2] + 1.75, b=coef(mat)[1], lwd=2,
       col="blue") #no tube, airport
abline(a=sum(coef(mat)[2:4]) + 1.75, b=coef(mat)[1], lwd=2,
       col="orange") #tube, airport
```
SFO outlier is Millbrae-SFO trip. Fare = 4.05

```{r}
# Residuals
# Airport-only model
par(mfrow=c(2, 2))
plot(mat)
par(mfrow=c(1, 1))

plot(y=mat$residuals, x=BART.fares$Miles,
      xlab = "Driving Distance Between Stations (Miles)",
     ylab="Residual", main="Residuals of Model")

# look at mean residual by station
boxplot(sapply(split(c(mat$residual, mat$residual), 
             c(BART.fares$To, BART.fares$From)), mean))
lapply(split(c(mat$residual, mat$residual), 
             c(BART.fares$To, BART.fares$From)), length)
# Millbrae has an average residual of +52 cents!

fareBART <- function(miles, SFO=FALSE) {
  # coefficients are from lm object 'ma'
  return (1.75 + 0.1045*miles + SFO*4.50)
}
```